---
title: "Assignment 5: Data Visualization"
author: "Tallulah Bowden"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
#load packages
library(tidyverse)
library(here)
library(cowplot)
library(ggplot2)
library(ggthemes)

#check WD
#getwd()
#here()

#read in datasets
Peter_Paul <- read.csv(
"Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", 
  stringsAsFactors = T)
NEON <- read.csv(
  "Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv",
  stringsAsFactors = T)
#2 
#check date classes
class(NEON$collectDate)
class(Peter_Paul$sampledate)

#change from factor to date
NEON$collectDate <- ymd(NEON$collectDate)
Peter_Paul$sampledate <- ymd(Peter_Paul$sampledate)

#check date classes again
class(NEON$collectDate)
class(Peter_Paul$sampledate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <- theme_base()  + 
  theme(
    line = element_line(
      color = "black",
      linewidth = 1),
    legend.background = element_rect(
      color = "grey",
      fill = "light blue"),
    lenged.title = element_text(
      color = "blue"),
    axis.title.x = element_text(
      color = "black"),
    axis.title.y = element_text(
      color = "black"),
    legend.position = "right",
    complete = T)
    
theme_set(my_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
phosphorus_phosphate <- Peter_Paul %>%
ggplot(
  mapping = 
    aes(x = po4,
        y = tp_ug,
        color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = lm)+
  xlim(0,45)+
  labs(x = expression("PO" [4])) +
  labs(color = "Lake Name")+
  labs(y = "Total Phosphorus")+
  labs(title = "Total Phosphorus by Phosphate for each Lake")

phosphorus_phosphate
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
Peter_Paul_month <- 
  mutate(Peter_Paul,
    month_num = month(sampledate), 
    month_name_abb = month(sampledate, label = TRUE), 
    month_name_full = month(sampledate, label = TRUE, abbr = FALSE)
  )

temp_plot <- Peter_Paul_month %>%
  ggplot(
    mapping = 
      aes(x = month_name_abb,
          y = temperature_C,
          color = lakename))+
  geom_boxplot()+
  labs(x = "Month") +
labs(y = "Temperature\n(C)") +
labs(color = "Lake Name") +
  theme(legend.position = "top",
        axis.title.x = element_blank(),
        legend.key.size = unit(0.5, "lines"),
        legend.title = element_text(size = 10),
        legend.title.position = "top",
        axis.title.y = element_text(angle = 90,
                                    margin =
                                      margin(r = 15)),
        plot.margin = unit(c(0, 0, 0, 0), "cm"))
  
  
  
#temp_plot

TP_plot <- Peter_Paul_month %>%
  ggplot(
    mapping = 
      aes(x = month_name_abb,
          y = tp_ug,
          color = lakename))+
  geom_boxplot()+
  labs(x = "Month")+
  labs(y = "Total\nPhosphorus")+
  labs(color = "Lake Name")+
  theme(axis.title.x = element_blank(),
        legend.position = "none",
        axis.title.y = element_text(angle = 90,
                                    margin =
                                      margin(r = 15)),
        plot.margin = unit(c(0, 0, 0, 0), "cm"))

#TP_plot
      
TN_plot <- Peter_Paul_month %>%
  ggplot(
    mapping = 
      aes(x = month_name_abb,
          y = tn_ug,
          color = lakename))+
  geom_boxplot()+
  labs(x = "Month")+
  labs(y = "Total\nNitrogen")+
  labs(color = "Lake Name")+
  theme(legend.position = "none",
        axis.title.y = element_text(angle = 90,
                                    margin =
                                      margin(r = 15)),
        plot.margin = unit(c(0, 0, 0, 0), "cm"))

#TN_plot

plot_grid(
  temp_plot, 
  TP_plot, 
  TN_plot, 
  nrow = 3, 
  align = 'h')
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature follows a curve with May and June having colder 
temperatures and July through September having hotter temperatures. Going into
October, the temperature begins to drop again. Paul lake seems to have slightly
higher temperatures at nearly all months except October and November
Total phosphorus also starts with lower values in May and increases very
slightly through September The interquartile ranges also increase over this 
time.Paul lake consistently has lower total phosphorus values.
Total nitrogen follows a similar trend to total phosphorus. Paul lake also has 
lower values than Peter lake.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
needle_plot <- NEON %>%
  filter(functionalGroup == "Needles")%>%
  ggplot(mapping = 
           aes(x = collectDate,
               y = dryMass,
               color = nlcdClass))+
  geom_point()+
  labs(x = "Date")+
  labs(y = "Needle Dry Mass")+
  labs(color = "NLCD Class")+
  scale_color_discrete(
    labels = c("Evergreen Forest", "Herbaceous Grassland", "Scrub Shrub")
  )+
  theme(legend.position = "bottom",
        legend.key.size = unit(1, "lines"),
        legend.title = element_text(size = 10),
        legend.title.position = "top")

needle_plot

#7
needle_faceted_plot <- NEON %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(
    mapping = 
      aes(x = collectDate,
          y = dryMass))+
  geom_point()+
  facet_wrap(vars(nlcdClass), 
             ncol = 3, 
             labeller = labeller(nlcdClass = 
                        c("evergreenForest" = "Evergreen Forest",
                        "grasslandHerbaceous" = "Herbaceous Grassland",
                        "shrubScrub" = "Scrub Shrub")))+
  labs(x = "Date")+
  labs(y = "Needle Dry Mass")
  
needle_faceted_plot

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think the faceted plot (7) is more effective in this analysis.
Although I like having plots with color, in the first plot it was hard to have
to go to the legend and find which color represented which NLCD class and then
go back to the plot and find which date it represented and then what dry mass it
was associated with. There were a lot of steps to get meaningful information
out of that plot and it would be even harder to look at trends. In the 
faceted plot, it is much easier to look at one of the facets and be able to tell
what is happening with needle dry mass over our collection time.