---
title: "7: Lab - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Answer questions on M5/A5
2. Answer questions on M6 - GLMs
3. Practice more application GLM to real datasets


## Set up
```{r, message = FALSE}
library(tidyverse)
library(agricolae)   
library(here)
here()

EPAair <- read.csv(here("Data/Processed_KEY/EPAair_O3_PM25_NC1819_Processed.csv"), stringsAsFactors = TRUE)
# Set date to date format
EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")

Litter <- read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)
# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```



## Visualization and interpretation challenge

Create three plots, each with appropriately formatted axes and legends. Choose a non-default color palette.

1. geom_density of ozone divided by year (distinguish between years by adding transparency to the geom_density layer).
2. geom_boxplot of ozone divided by year. Add letters representing a significant difference between 2018 and 2019 (hint: stat_summary). 
3. geom_violin of ozone divided by year, with the 0.5 quantile marked as a horizontal line. Add letters representing a significant difference between 2018 and 2019. 

```{r}
#Exercise 1:
#plot 1
density_plot <- EPAair %>%
  ggplot(aes(Ozone, fill = factor(Year)))+
  geom_density(alpha=0.5)+
  labs(color = "Year")
density_plot

#plot2
boxplot_plot <- EPAair %>%
  ggplot(mapping = 
           aes(x = factor(Year), y = Ozone))+
  geom_boxplot()+
  labs(x = "Year")+
  stat_summary(geom = "text", 
               fun = max, 
               vjust = -1, 
               size = 4,
               label = c("a", "b"))
boxplot_plot

#plot3
violin_plot <- EPAair %>%
  ggplot(mapping = 
           aes(x = factor(Year), y = Ozone))+
  geom_violin(draw_quantiles = 0.5)+
  labs(x="Year")+
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("a", "b"))
violin_plot

```

## Linear Regression

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

For the NTL-LTER dataset, can we predict PM2.5 from Ozone?

```{r}

#Exercise 2: Run a linear regression PM2.5 by Ozone. Find the p-value and R-squared value. 
lm1 <- lm(data = EPAair, 
          PM2.5 ~ Ozone)
summary(lm1)


#Exercise 3: Build a scatterplot. Add a line and standard error for the linear regression. Add the regression equation to the plot
lm1_plot <- 
  ggplot(EPAair, aes(x = Ozone, y = PM2.5))+
  geom_point()+
  geom_smooth(method = "lm")+
  geom_text(x = 100,
            y = 75,
            label = expression("PM2.5 = 0.38*Ozone + 15.64"))
lm1_plot

```


## AIC to select variables

What other variables can we add to improve model?

```{r}

#Exercise 4: Build correlation plots and identify more possible explanatory variables to add to the regression.
library(corrplot)
EPAselected <- 
  EPAair %>%
  select(mean_Lat:Ozone) %>%
  na.omit()
EPAcorr <- cor(EPAselected)
corrplot(EPAcorr, method = "ellipse")
corrplot.mixed(EPAcorr, upper = "ellipse")



#Exercise 5: Choose a model by AIC in a Stepwise Algorithm. Do the results from AIC match the variables you selected on Exercise 4?
EPA.naomit <- EPAair %>%
  na.omit()

EPAaic <- lm(data = EPA.naomit, PM2.5 ~ mean_Lat + Ozone + 
              mean_Lng + Month + Year)

#Choose a model by AIC in a Stepwise Algorithm
step(EPAaic)




#Exercise 6: Run another regression using the variables selected on Exercise 6. Compare r-squared value with the one from Exercise 2. 
TPmodel <- lm(data = EPA.naomit, PM2.5 ~ mean_Lat + Ozone + mean_Lng + Month)
summary(TPmodel)
summary(lm1)
#R-squared improved from 0.1314 to 0.192

```



## Litter Exercise

```{r}
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))

# Format ANOVA as aov
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova)

# Extract groupings for pairwise relationships
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE)
Litter.Totals.groups$groups

Litter.Totals <- Litter.Totals %>%
  mutate( treatgroups = Litter.Totals.groups$groups[plotID,2])

# Graph the results
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)

```

```{r}
#Exercise 7: Improve the plot
# a. order by groups from high to low mean
assign.group <- function(site)
return(Litter.Totals.groups$groups[as.character(site),2])

treatgroups=array(0,length(Litter.Totals))
for( i in length(Litter.Totals)){
treatgroups = assign.group(Litter.Totals$plotID)
}

Litter.Totals <- cbind(Litter.Totals,groups = treatgroups)
Litter.Totals$plotID <- reorder(Litter.Totals$plotID,Litter.Totals$dryMass,FUN=mean)

Litter.Totals.plot2 <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
geom_boxplot() +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
label = Litter.Totals.groups$groups[,2]) +
labs(x = "Plot ID", y = "Dry Mass (g)") +
ylim(0, 85)
print(Litter.Totals.plot2)
```
